Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-15 main.tex
Explain @ symbols in DLMF macros
Implement several requested minor changes

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. One example of such a macro is given in Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occurs for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{}\footnote{The mention of
specific products, trademarks, or brand
names is for purposes of identification only. Such mention is not to be interpreted
in any way as an endorsement or certification of such products or brands by the
National Institute of Standards and Technology, nor does it imply that the products
so identified are necessarily the best available for the purpose. All trademarks
mentioned herein belong to their respective owners.} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition for what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

Another approach tries to avoid translation problems by allowing computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the abilities of the compiler and therefore are not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}.

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
	\mathscr{W} \left\{ w_1(z), w_2(z) \right\} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitely define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some types together for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table}

@@

If we take a closer look to mathematical notations we discover many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

An exponent will be translated and shifted to the end of the translated semantic macro, 
since one usually writes the exponent after the arguments of a function in \gls*{cas}. Hence, the function translates and stores the exponent in Line~\ref{line:next_caret}. One could ask what happens when there is an exponent given before and after the arguments. The \verb|MacroTranslator| only translates the following siblings until each argument is translated. The first exponent will be shifted to the end. If right after the translated macro (with all arguments) follows another exponent, we interpret it as another exponent for the whole previous expression. In that case, it would be the macro with the first translated exponent. Table~\ref{tab:multi-expo} shows an example for the trigonometric cosine function with multiple exponents. 
\begin{table}[ht]
\centering
\begin{tabular}{lc}
\hline
Displayed As & \rule{0pt}{0.9\normalbaselineskip} $\cos^2(x)^2$ \\
Semantic \LaTeX & \verb|\cos^2@{x}^2| \\
Translated \Maple{} Expression & \verb|((cos(x))^(2))^2|\\
\hline
\end{tabular}
\caption{A trigonometric cosine function example with exponents before and after the argument.}
\label{tab:multi-expo}
\end{table}

As you can see, the input expression in semantic \LaTeX{} is interpreted as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^2 \right)^2.
\end{equation}

>>>>

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. The macros may also contain $@$ symbols preceding the variables of the function. The number of the symbols is used to switch between different notation styles, e.g., $\cos(x)$ and $\cos x$. One example of such a macro is given in Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occurs for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continuous function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{}\footnote{The mention of
specific products, trademarks, or brand
names is for purposes of identification only. Such mention is not to be interpreted
in any way as an endorsement or certification of such products or brands by the
National Institute of Standards and Technology, nor does it imply that the products
so identified are necessarily the best available for the purpose. All trademarks
mentioned herein belong to their respective owners.} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition for what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}.

Another approach tries to avoid translation problems by allowing computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the abilities of the compiler and therefore are not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}, but it is possible to integrate \gls*{cas} expression directly into \TeX{} documents.

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
	\mathscr{W} \left\{ w_1(z), w_2(z) \right\} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. Note that normal parentheses do not create balanced expressions. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$). \\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some types together for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table}

@@

If we take a closer look to mathematical notations we discover many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{prefix notation}, \textit{Warsaw Notation} or \textit{{\L}ukasiewicz notation}. It was invented by J. {\L}ukasiewicz 1924 to create a parenthesis-free notation~\parencite{Hamblin1962}. Note that this notation is indeed parenthesis-free as long as all operators have the same arity.} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}. Also invented by J. {\L}ukasiewicz. Same as \gls*{pol} it does not need parenthesis as long as all operators have the same arity.} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

An exponent will be translated and shifted to the end of the translated semantic macro, 
since one usually writes the exponent after the arguments of a function in \gls*{cas}. Hence, the function translates and stores the exponent in Line~\ref{line:next_caret}. One could ask what happens when there is an exponent given before and after the arguments. The \verb|MacroTranslator| only translates the following siblings until each argument is translated. The first exponent will be shifted to the end. If right after the translated macro (with all arguments) follows another exponent, we interpret it as another exponent for the whole previous expression. In that case, it would be the macro with the first translated exponent. Table~\ref{tab:multi-expo} shows an example for the trigonometric cosine function with multiple exponents. As illustrated in Table~\ref{tab:multi-expo}, the translator interprets the input as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^n \right)^m.
\end{equation}

\begin{table}[ht]
\centering
\begin{tabular}{lc}
\hline
Displayed As & \rule{0pt}{0.9\normalbaselineskip} $\cos^n(x)^m$ \\
Semantic \LaTeX & \verb|\cos^n@{x}^m| \\
Translated \Maple{} Expression & \verb|((cos(x))^(n))^m|\\
\hline
\end{tabular}
\caption{A trigonometric cosine function example with exponents before and after the argument.}
\label{tab:multi-expo}
\end{table}
